Verification of the CVM algorithm with a New Recursive Analysis Technique

### 2025-02-06 paper/sec/intro.tex
Paper: Edit intro a bit more.
Avoiding reference to the simplicity of the original proof, which contradicts the main point of the paper.
Keep most of the changes previous commit.
Adding summary of sections.

In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem which was very unexpected in the community~\cite{quanta}.
Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
One striking property of the CVM algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.

Another property of CVM is its simplicity, both in terms of its description---the algorithm is listed in its entirety in~\cref{alg:cvm}---and its pen-and-paper correctness proof, which only requires undergraduate-level exposure to randomized algorithms.
This is why its authors called it ``an algorithm for the text book''.
The motivation of this work is to explore whether CVM's simplicity transfers into a mechanization within the Isabelle proof assistant~\cite{nipkow2002}.

@@

Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
Briefly, the pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
Yet, for~\cref{alg:cvm} the random operations are far from being independent.\footnote{This was an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022}; a later version by the same authors~\cite{DBLP:journals/corr/abs-2301-10191} provides a correct proof.
The original error serves as a side motivation for this work.}
For example, in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations; similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
The aforementioned sequence of transformations bounds the error of~\cref{alg:cvm} in terms of another randomized algorithm with more desirable independence properties.
Indeed, it seems impossible to analyze~\cref{alg:cvm} directly using known methods.

This is the point where our new analysis technique comes in---it is very similar to how deterministic algorithms are verified using a loop invariant.
The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
\[
  \expect [ h ] \leq h(c)
\]
where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
By first establishing such an invariant for~\cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a simple but substantially more general proof of the CVM algorithm.
We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to interactive theorem proving.

@@

Verification of CVM using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.
We detail some of the challenges faced when mechanizing the reduction arguments used by Chakraborty et al.~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191} in Section~\todo{x}.

The formalized CVM variant showcases the utility of our new proof technique.
In this variant, the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-sized subset of $\chi$, instead of independently deciding for each element, whether to keep it.
This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in line 11 becomes obsolete.
More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected result is exactly the cardinality of the elements in the stream; this is a new property, that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.

@@

% TODO: move it into the discussion of the variant, I think it is better once the algorithm is shown
%This is because indicator functions of $n$-subsets form negatively associated random variables (RV), even though they are not independent, with which we can modify the subsampling step to the form we described above.

@@

\todo{Classic summary of sections.}

>>>>

In 2022, Chakraborty, Vinodchandran, and Meel~\cite{chakraborty2022} published a marvelous streaming algorithm for the distinct elements problem, which was very unexpected in the community~\cite{quanta}.
Indeed, Knuth later wrote a note on the algorithm~\cite{knuthnote}, pointing out its interesting properties and christening it the \emph{CVM} algorithm (which we use for the rest of this paper).
One striking property of the CVM algorithm is that, in contrast to every other known algorithm for the problem, it does not rely on hashing the stream elements.
Instead, the algorithm could theoretically be implemented in a setting where objects in the data stream only allow for equality comparisons.
Another property is its simplicity---which is why the authors called it ``an algorithm for the text book''---the algorithm is shown in its entirety in~\cref{alg:cvm}.

@@

%Curiously, our formalization takes a substantially different route compared to the original proof~\cite{chakraborty2022,DBLP:journals/corr/abs-2301-10191}.
Briefly, the pen-and-paper analysis for CVM relies on a sequence of transformations of the algorithm.
The reason for these transformations is that standard methods for analyzing randomized algorithms, such as Chernoff--Hoeffding bounds, usually make statements about independent random variables.
Yet, for \cref{alg:cvm} the random operations are far from being independent.\footnote{This was an incorrect claim in the initial published proof of CVM~\cite[Claim 6]{chakraborty2022}; a later version by the same authors~\cite{chakraborty2023} provides a correct proof.
The original error serves as a side motivation for this work.}
For example, in line 3 the Bernoulli distribution is sampled for a value $p$, which itself depends on previous random operations; similarly, the subsampling step in line 9 is only applied if the buffer is full, which also depends on previous random operations.
The aforementioned sequence of transformations results in another randomized algorithm which can be analyzed using standard methods, from which the desired results about the original algorithm can be deduced.
Indeed, it seems impossible to analyze~\cref{alg:cvm} directly using known methods.

In this paper, we present a new technique for analyzing randomized algorithms.
The technique yields a direct and substantially more general proof of the CVM algorithm.
It is very similar to how deterministic algorithms are verified using a loop invariant.
The key difference is that our choice of ``loop invariant'' for the randomized streaming algorithm is a functional probabilistic inequality, namely, we consider invariants of the form:
\[
  \expect [ h ] \leq h(c)
\]
where $h$ is allowed to range over a class of functions, and the expectation is taken over the distribution of the state of the algorithm after consuming each stream element.
By first establishing such an invariant for~\cref{alg:cvm}, we can then use it (via different choices of $h$) to establish error bounds for the algorithm.
For the rest of this paper, we explain this technique, its mechanization, and show how it leads to a simple but substantially more general proof of the CVM algorithm.
We believe the new proof remains accessible at the undergraduate level, albeit with some exposure to interactive theorem proving.

@@

The verification of the CVM algorithm using our new technique requires only 1044 lines in Isabelle~\cite{nipkow2002}, while the original proof required \todo{x} lines.
We detail some of the challenges faced when mechanizing the reduction arguments used by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023} in \cref{sec:transformation_based_proof}.

To show that our new technique is more general, we introduce a new variant of the CVM algorithm, where the subsampling step in line 9 of \cref{alg:cvm} selects a random $m$-subset of $\chi$, instead of independently deciding for each element, whether to keep it.
This variant has the benefit that it is \emph{total} (never returns $\bot$) because the second check in line 11 becomes obsolete.
More interestingly, the resulting variant is \emph{unbiased}, i.e., the expected result is exactly the cardinality of the elements in the stream; this is a new property, that neither the original CVM algorithm nor classic algorithms for the distinct elements problem possess.

@@

The following section provides more background information, in particular, about the semantics of randomized algorithms.
After which, we introduce our new technique, showing how probabilistic loop invariants can be used to establish tail bounds for the original CVM algorithm.
\cref{sec:negdep} introduces the concept of negative association and our new total and unbiased variant of the CVM algorithm.
\cref{sec:formalization} and \cref{sec:formalization_neg_dep} present details about the formal verification of both variants of the algorithm, as well as, our new library on negatively associated random variables.
In \cref{sec:transformation_based_proof}, we present details about our alternative verification of the CVM using the transformation based proof by Chakraborty et al.
The last sections conclude with a brief overview of related work and a summary of our results.
% TODO: move it into the discussion of the variant, I think it is better once the algorithm is shown
%This is because indicator functions of $n$-subsets form negatively associated random variables (RV), even though they are not independent, with which we can modify the subsampling step to the form we described above.

@@

### 2025-02-06 paper/sec/original_proof.tex
Paper: Edit intro a bit more.
Avoiding reference to the simplicity of the original proof, which contradicts the main point of the paper.
Keep most of the changes previous commit.
Adding summary of sections.

\section{Original Proof}
In this section we describe the interesting parts of the original proof by Chakraborty et al.~\cite{chakraborty2022}.
This highlights in part, why it requires a lot of work to formalize, and why we were motivated to develop our new approach using probabilistic invariants.

>>>>

\section{Transformation-based Proof}\label{sec:transformation_based_proof}
In this section we describe the interesting parts of the original transformation-based proof by Chakraborty et al.~\cite{chakraborty2022}.
This highlights in part, why it requires a lot of work to formalize, and why we were motivated to develop our new approach using probabilistic invariants.
